---
title: "6.1 For Loops"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

## 6.1.1: Introduction to For Loops

---

### Repetitive Tasks

Often in life we need to perform the same task multiple times, each time performing the task for a different item:
- For each dish in the sink, wash the dish
- For each course you are taking, add the exam date to your calendar

These situations also show up often in biology. For example, let’s say you are working in the lab and have six samples. For each sample you need to add 10 uL of lysis buffer and then place it on ice.

It is a repetitive task, in which the the sample you are working with changes but the procedure does not change.

If we were to write out a to-do list for this task it could look like this:

    - Add 10 uL of lysis buffer to sample 1
    - Place sample 1 on ice
    - Add 10 uL of lysis buffer to sample 2
    - Place sample 2 on ice
    - Add 10 uL of lysis buffer to sample 3
    - Place sample 3 on ice
    - Add 10 uL of lysis buffer to sample 4
    - Place sample 4 on ice
    - Add 10 uL of lysis buffer to sample 5
    - Place sample 5 on ice
    - Add 10 uL of lysis buffer to sample 6
    - Place sample 6 on ice

Or we could streamline the to-do list:

    For X in sample 1, sample 2, sample 3, sample 4, sample 5, sample 6:
      - Add 10 uL of lysis buffer to X
      - Place X on ice

These two to-do lists are exactly the same--both outline the same set of tasks to be performed--but by using X to represent each individual sample, we can save ourselves a lot of excess writing. 
      
---

### For Loops

In computer programming, *loops* perform repetitive tasks. A *for loop* performs the same programming task for each item in a set of items:

    for (i in items) {
      ... code block ...
    }

This is equivalent to saying: "for each item i in the set of items, execute this code block"

The syntax of a for loop will look similar to the syntax for conditional statements. As with conditional statements, the code within the curly braces is preceded by a tab for greater clarity when reading code. 

To better understand how a for loop works, let's walk through an example. Let's say we want to take the natural log of the numbers 1, 3, and 5. We could write the following 3 lines of code:

```{r}
log(1)
log(3)
log(5)
```

We could also assign each number to a variable before taking the logarithm:

```{r}
x <- 1
log(x)
x <- 3
log(x)
x <- 5
log(x)
```

When we write the code like this, we are taking each item (1, 3, and 5) and running the exact same piece of code `log(x)`. This indicates that we can convert this code into a for loop:

    for (i in items) {
      ... code block ...
    }

```{r}
for (x in c(1,3,5)) {
  print(log(x))
}
```

**Note that like conditional statements, we need to use the print function inside the curly braces**

Internally, this for loop runs the exact same code as the previous code block. A for loop will run once for each item in the set of items (in this for loop, that's three times). Each time it runs for an item it is called an *iteration* of the loop. In the first iteration, the variable `x` will take on the value of the first item in the set (`1`), and then the code inside will be executed In the second iteration, `x` will take on the value of the second item in the set (`3`), and then the code inside will be executed. Finally, `x` will take on the value of the third item in the set (`5`) and the code inside will be executed.

For loops save you time and space when writing code, as you don't have to write the same code over and over again.

---

## 6.1.2: For Loop Techniques

---

### Modifying Data Structures

For loops can be used to alter existing data structures, including:
- Adding elements to vectors
- Adding rows to matrices or data frames
- Editing each element in a vector
- Editing each value in a data frame column
Etc.

Let's look at an example of adding elements to a vector. First let's create an empty vector called "celsius":

```{r}
celsius <- c()
celsius
```

Empty vectors output `NULL` as they do not have a value. 

Now we can use a for loop to add elements to the vector. The for loop below takes a vector that contains three temperatures in Kelvin as its set of items. At each iteration of the loop, the current temperature is assigned to the variable `temp`. 

This first line of code within the curly braces converts the temperature from Kelvin to Celsius by subtracting 273.15 degrees. The second line adds the new value to end of the `celsius` vector with the combine function, and the last line outputs the current value of the `celsius` vector.

```{r}
for (temp in c(292.48, 301.29, 287.32)) {
  t_C <- temp - 273.15
  celsius <- c(celsius, t_C)
  print(celsius)
}
```

When we run the code, we can see that three lines are printed, one for each iteration of the loop. At the end of the first iteration, one value has been added to `celsius`, two have been added at the end of the second iteration, and three at the end of the third.

---

### Looping Through Indices

Often it is helpful to loop through the positional indices of a vector so we can access the index of each element in each iteration of the for loop.

To loop through positional indices we can create a sequence from 1 to the length of the vector: `1:length(vector)`.

For example, let's say we have a vector of weights (in grams) for four mice and we want to create a vector of sample names for the mice:

```{r}
mouse_wts <- c(23.7, 21.4, 26.8, 28.3)
mouse_samples <- c()
```

The `paste()` function takes character values and a separator value to combine multiple character values into value, each separated by the separator value. For example, to add together the words "sample", "A", and "control" separated by dashes we can use the `paste()` function: 

```{r}
paste("sample", "A", "control", sep="-")
```

We can loop through each positional index in mouse weights, and use the `paste()` function to paste together the word "mouse" and the current index with an underscore in between. Then we can add this sample name to our `mouse_samples` vector:

```{r}
for (i in 1:length(mouse_wts)) {
  sample <- paste("mouse", i, sep="_")
  mouse_samples <- c(mouse_samples, sample)
}
mouse_samples
```

When we look at the value of `mouse_samples`, we can see that we have a vector with the names "mouse_1", "mouse_2", "mouse_3", and "mouse_4".

Another reason to loop through the positional indices of a vector is to be able to access both the index AND the value of each element.

For example, if we want to go through each of the elements in this vector of mouse samples and edit (replace) each value based on the current value, we need to be able to access value as well as the index of the value. The index allows us to replace the value at the correct index.If we only know an element value, we cannot access the index, but if we know the positional index of an element we can also access the value. 

For example, to replace each sample name with a shortened version by replacing the letters "ouse" with "", we need the value of the sample name for replacement, and the index of the sample name to assign the new name.

```{r}
for (i in 1:length(mouse_samples)) {
  sample <- mouse_samples[i]
  short_name <- gsub("ouse", "", sample)
  mouse_samples[i] <- short_name
}
mouse_samples
```

At the start of the above loop we retrieve the value at the index, so that in each iteration, `i` contains the index and `sample` contains the value at that index. With `gsub()` we can replace the letters "ouse" in `sample` with nothing, shortening the sample name. Finally, we can use the index again to replace the value in `mouse_samples` at the appropriate index with the new shortened name. 

Using the `mouse_samples` vector, names can be added to the `mouse_wts` vector:

```{r}
names(mouse_wts) <- mouse_samples
mouse_wts
```

This vector can also be looped through using the names of the vector. For example, to change each mouse weight in the vector from grams to pounds, the name indices can be looped through:

```{r}
for (sample in names(mouse_wts)) {
  wt_pounds <- mouse_wts[sample] * 0.00220462
  mouse_wts[sample] <- wt_pounds
}
mouse_wts
```

---

## 6.1.3: For Loop Applications

---

### Expression Correlation

Now we will examine some practical examples of for loops in biological research. We will be using brain tissue expression data sourced from the Human Protein Atlas (https://www.proteinatlas.org/humanproteome/brain). 

```{r}
brain_expression <- read.csv("/home/jovyan/Week.6/6.1.For.Loops/brain_expression.csv", stringsAsFactors = FALSE)
brain_expression
```

Each of the rows represent one of 15,271 genes. The first column is the Ensembl Gene ID of each gene, which is a unique gene identifier. The second to eleventh columns contain normalized gene expression values for 10 different brain tissues. Let's save the names of these tissues (which are the column names for columns 2-11) in a vector named `brain_tissues`:

```{r}
brain_tissues <- colnames(brain_expression)[2:11]
brain_tissues
```

To calculate the correlation (R) of tissue expression in the amygdala and the thalamus, the `cor()` function can be used. 

```{r}
r <- cor(brain_expression[, "amygdala"], brain_expression[, "thalamus"])
r <- round(r, 2)
paste("amygdala and thalamus:", r)
```

To output the correlation in a readable format, the correlation was also rounded and the `paste()` function was used. Note that the default value for the `sep` argument in the `paste()`  function is a single space " ".

If we want to calculate the correlation in gene expression between the amygdala and each of the other brain tissues (including itself), we could write out the above code 10 times for a total of 30 lines of code.

Alternatively, we can loop through each tissue using a for loop and calculate the correlation between the tissue in the current iteration and the amygdala.

For this loop, our items will be each of the elements the `brain_tissues` vector. In each iteration, we can determine the correlation between the amygdala and the iteration tissue by retrieving the values in both columns and using them as arguments to the `cor()` function. Then the resulting correlation is rounded and output using the `print()` and `paste()` functions to provide information about which tissues are being compared:

```{r}
for (tissue in brain_tissues) {
  r <- cor(brain_expression[, "amygdala"], brain_expression[, tissue])
  r <- round(r, 2)
  print(paste("amygdala and", tissue, ":", r))
}
```

This is much easier than writing the 30 lines of code that would be needed otherwise.

---

### Looping Through Rows

So far we have been writing loops to iterate through vectors or vector indices. It is often also helpful to loop through data frame row indices. This allows you to access all of the values in a particular row within the loop.

The hindbrain is composed of the pons, the medulla oblongata, and the cerebellum. In the `brain_expression` data frame, there is one column for `cerebellum` and one for `pons.and.medulla`. If we wanted to know the maximum expression of each gene in any component of the hindbrain, we could add a new column to contain the maximum expression of each gene. To do this we will need to determine the highest expression of the "cerebellum" column value and the "pons.and.medulla" column value.

First, we can add a new column called "hindbrain.max" to the `brain_expression` data frame with all values equal to `NA`. The for loop iterates through each row index in `brain_expression`, retrieves the maximum value for that row between the two columns, and assigns the higher value to the `hindbrain.max` column for the row currently being iterated:

```{r}
brain_expression$hindbrain.max <- NA
for (row in 1:nrow(brain_expression)) {
  hindbrain.row.max <- max(brain_expression[row, c("pons.and.medulla", "cerebellum")])
  brain_expression$hindbrain.max[row] <- hindbrain.row.max
}
brain_expression
```

---